Effects of Different Pain Relievers and Dosages on Cognitive Retention

### 2025-04-01 paper/paper.qmd
feedback fixes, streamlined tables, equations fixed x3

## Research Questions
1. *How does the type of pain reliever (Aspirin 500 mg, Paracetamol 500 mg, Tramadol 50 mg, and Placebo) affect memory game performance?*
- Null Hypothesis: All pain relievers at normal dosages act the same for cognitive retention game performances.
- Alternative Hypothesis: There exists at least one pair of distinct pain relievers at normal dosages which act differently for cognitive retention game performances.

2. *Is there an interaction effect between the type of pain reliever and dosage level on the improvement in memory game performance?*
- Null Hypothesis: There is no interaction effect between the type of pain reliever and dosage level on the improvement in memory task performance.
- Alternative Hypothesis: There is an interaction effect between at least one of the types of pain reliever and dosage level on the improvement in memory task performance.

@@

## Study Design and Treatments   
We selected three brands of painkillers: Aspirin, Paracetamol, and Tramadol. Additionally, we considered a placebo treatment. As a result, this study involves 2 factors: pain reliever type (4 levels - Aspirin, Paracetamol, Tramadol, Placebo) and dosage (2 levels - Low, High). Crossing the two factors provides us with a total of *8 treatment groups*, with **30 participants** assigned to each. The groups are as follows:

@@

### Nuisance Variable: Age  
To reduce bias, we consider **controlling** the experiment by limiting participant ages to be strictly 18+. Hence in this study, **age** acts as a **nuisance variable**.
We also used **blocking** for age by ensuring participants were categorized into three even groups to prevent bias created through age. @tbl-summary-5 shows the breakdown of the groups.

@@

1. Quantitative variables: Memory Game Scores, Memory Test Cards Scores 
2. Qualitative variables: Type of pain reliever, dosage level (low/high)

@@

```{r}
#| label: tbl-summary-5
#| tbl-cap: "Summary of Age Groups"
#| echo: false
#| warning: false
#| message: false

age_counts <- table(clean_data$age_group)

# summary dataframe
age_summary <- data.frame(
  `Age Group` = names(age_counts),
  Count = as.numeric(age_counts),
  Percentage = round(as.numeric(age_percentages), 2)
)

colnames(age_summary)[colnames(age_summary) == "Age.Group"] <- "Age Group"

@@

@tbl-summary-1 shows the summary statistics of memory game scores before and after the treatment. The mean is similar across all groups, however, the standard deviation (SD) is relatively high. The median game scores remain close to the mean, and the IQR (Interquartile Range) is smaller than the SD. The cards memory task shows little change before and after treatment, suggesting that pain relievers may not have a major effect on this specific task.

@@

```{r}
#| label: tbl-summary-3
#| tbl-cap: "Summary of Dosage Levels"
#| echo: false
#| warning: false
#| message: false

@@

@tbl-summary-3 shows the dosage levels assigned to participants. There is an equal split between high-dose (120 participants, 50\%) and low-dose (120 participants, 50\%) conditions.

@@

$$
\text{Y}_i = \beta_0 + \beta_1 \cdot \text{drug}_i + \beta_2 \cdot \text{dosage\_level}_i + \beta_3 \cdot (\text{drug}_i \times \text{dosage\_level}_i) + \varepsilon_i
$$

@@

- $\beta_0$ is the intercept.

- $\beta_1$, $\beta_2$, and $\beta_3$ are coefficients for the predictors and interaction term.

- $\varepsilon_i$ is the random error term.

@@

kable(bartlett_table, format = "latex", booktabs = TRUE)
```
# Analysis
The method of two-way ANOVA is used to estimate the mean of a memory game score changes according to the levels of two categorical variables, being drug type and dosage level. Our research question explores if there is an interaction effect between at least one of the types of pain reliever and dosage level on the improvement in memory task performance.

We verified the assumption of the two-way ANOVA model as seen from graph normality, using Q-Q plot as seen in @fig-normal-plot. The Shapiro-Wilk Test confirms a p-value = 0.05553 as per @tbl-shapiro, so we fail to reject the null hypothesis and the result confirms that the data-set is normally distributed.  
Our second assumption of Homogeneity of Variance is verified by the p-value = 0.3346 from the Bartlett test in @tbl-bartlett, which is greater than 0.05. Due to this, we fail to reject the null hypothesis.
Thirdly, for our two-way ANOVA model, the different combinations of pain reliever type and dosage levels were randomly assigned to all participants, in random order.

@@

# Discussion
## Interpretation of Results  
The two-way ANOVA analysis revealed no statistically significant main effects of drug type (p = 0.25) or dosage level (p = 0.26) on memory game scores. Similarly, the interaction effect between drug type and dosage level was not significant (p = 0.24). These results suggest that neither the type of pain reliever (Aspirin, Paracetamol, Tramadol, or Placebo) nor its dosage (low vs. high) had a meaningful impact on memory game scores, which is representative of cognitive improvement in this study. The lack of significance persisted even after accounting for age as a confounding variable, which was controlled through participant stratification into three balanced groups (18–34, 35–50, and 50+ years).

@@

# Conclusion
This study found no conclusive evidence that pain relievers or their dosages significantly affect cognitive retention. However, the exploratory trends observed, particularly with Paracetamol and Tramadol, highlight the need for larger-scale studies to investigate subtle or context-dependent effects. Future research could incorporate longitudinal designs, broader age-specific analyses, and alternative cognitive metrics to deepen understanding of painkillers’ cognitive implications.

>>>>

## Research Questions
1. *How does the type of drug (Aspirin 500 mg, Paracetamol 500 mg, Tramadol 50 mg, and Placebo) affect memory game performance?*
- Null Hypothesis: All drug types at normal dosages act the same for cognitive retention game performances.
- Alternative Hypothesis: There exists at least one pair of distinct drug types at normal dosages which act differently for cognitive retention game performances.

2. *Is there an interaction effect between the type of pain reliever and dosage level on the improvement in memory game performance?*
- Null Hypothesis: There is no interaction effect between the type of drug and dosage level on the improvement in memory task performance.
- Alternative Hypothesis: There is an interaction effect between at least one of the types of drug and dosage level on the improvement in memory task performance.

@@

## Study Design and Treatments   
We selected three brands of painkillers: Aspirin, Paracetamol, and Tramadol. Additionally, we considered a placebo treatment. As a result, this study involves 2 factors: drug type (4 levels - Aspirin, Paracetamol, Tramadol, Placebo) and dosage (2 levels - Low, High). Crossing the two factors provides us with a total of *8 treatment groups*, with **30 participants** assigned to each. The groups are as follows:

@@

### Nuisance Variable: Age  
To reduce bias, we consider **controlling** the experiment by limiting participant ages to be strictly 18+. Hence in this study, **age** acts as a **nuisance variable**.
We also used **blocking** for age by ensuring participants were categorized into three even groups to prevent bias created through age.These groups are 18-34, 35-50, and 50+.

@@

1. Quantitative variables: Memory Game Scores, Memory Test Cards Scores 
2. Qualitative variables: Type of drug, dosage level (low/high)

@@

@@

@tbl-summary-1 shows the summary statistics of memory game scores before and after the treatment. The mean is similar across all groups, however, the standard deviation (SD) is relatively high. The median game scores remain close to the mean, and the IQR (Interquartile Range) is smaller than the SD. The cards memory task shows little change before and after treatment, suggesting that drug types may not have a major effect on this specific task.

@@

@@

@@

$$
\text{Y}_i = \beta_0 + \beta_1 \cdot \text{X}_i + \beta_2 \cdot \text{Z}_i + \beta_3 \cdot (\text{X}_i \times \text{Z}_i) + \varepsilon_i
$$

@@

- $\beta_0$ is the intercept.
- $X_i$ is the dummy variable for drug type.
- $Z_i$ is the dummy variable for dosage level.
- $\beta_1$, $\beta_2$, and $\beta_3$ are coefficients for the predictors and interaction term.
- $\varepsilon_i$ is the random error term.

@@

kable(bartlett_table, format = "latex", booktabs = TRUE)
```
# Analysis
The method of two-way ANOVA is used to estimate the mean of a memory game score changes according to the levels of two categorical variables, being drug type and dosage level. Our research question explores if there is an interaction effect between at least one of the types of drug types and dosage level on the improvement in memory task performance.

We verified the assumption of the two-way ANOVA model as seen from graph normality, using Q-Q plot as seen in @fig-normal-plot. The Shapiro-Wilk Test confirms a p-value = 0.05553 as per @tbl-shapiro, so we fail to reject the null hypothesis and the result confirms that the data-set is normally distributed.  
Our second assumption of Homogeneity of Variance is verified by the p-value = 0.3346 from the Bartlett test in @tbl-bartlett, which is greater than 0.05. Due to this, we fail to reject the null hypothesis.
Thirdly, for our two-way ANOVA model, the different combinations of drug type and dosage levels were randomly assigned to all participants, in random order.

@@

# Discussion
## Interpretation of Results  
The two-way ANOVA analysis revealed no statistically significant main effects of drug type (p = 0.25) or dosage level (p = 0.26) on memory game scores. Similarly, the interaction effect between drug type and dosage level was not significant (p = 0.24). These results suggest that neither the type of drug (Aspirin, Paracetamol, Tramadol, or Placebo) nor its dosage (low vs. high) had a meaningful impact on memory game scores, which is representative of cognitive improvement in this study. The lack of significance persisted even after accounting for age as a confounding variable, which was controlled through participant stratification into three balanced groups (18–34, 35–50, and 50+ years).

@@

# Conclusion
This study found no conclusive evidence that drug types or their dosages significantly affect cognitive retention. However, the exploratory trends observed, particularly with Paracetamol and Tramadol, highlight the need for larger-scale studies to investigate subtle or context-dependent effects. Future research could incorporate longitudinal designs, broader age-specific analyses, and alternative cognitive metrics to deepen understanding of painkillers’ cognitive implications.
